Temporal ordering and registration of images in studies of developmental dynamics

### 2015-02-11 image_analysis_paper_development2/manuscript.tex
added yannis/amit comments

Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
%
A schematic illustration of angular synchronization is shown in \fig~\ref{fig:schematics}A, where each image is represented as a vector, and the goal is to align the entire set of vectors given pairwise alignment measurements.
%
We first compute the angles needed to align pairs of vectors (or images), which in general requires no notion of a template function \citep{ahuja2007template}.
%
In this work, we aligned pairs of images with respect to rotations by exhaustively searching over a discretized space of rotation angles.
%
However, pairwise alignments can be computed by aligning appropriate image landmarks or features \citep{ian1998statistical}.
%
When the data are noisy, these pairwise measurements may by inaccurate, and so we utilize {\em all} pairwise measurements to align the set of images robustly.
%
Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see \SI); this is illustrated in \fig~\ref{fig:schematics}B.
%
In this schematic, registration via angular synchronization is trivial, as the pairwise measurements contain no noise.
%
However, the algorithm can register data sets even when many of the pairwise measurements are inaccurate \citep{singer2011angular}.

@@

To assess the robustness of the proposed methodology, we repeated this procedure with four additional data sets extracted from independent live imaging studies spanning the same developmental time period.
%
The results are shown in \fig~\ref{fig:drosophila_live_imaging}E.
%
The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than 95\%, indicating that our methodology can reproducibly order data of this type.
%
In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
%
At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \citep{singer2011angular}.
%
%Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
%%
%To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%%
%Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered.
%%
%\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%%
%As expected, the accuracy in the recovered ordering increases with the number of images.
%%
%At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
%%
%Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignment, and the high temporal resolution of our live imaging study allows us to order the images accurately.
%
%In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

@@

As another validation for the proposed methodology, we applied our algorithm to a time-lapse movie  of zebrafish embryogenesis.
%
We used a publicly available live imaging data of zebrafish embryogenesis (\texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov}, \citet{karlstrom1996flipbook}).
%
Taken with a differential interference contrast (DIC) microscope, the movie records the first $17$~hours of zebrafish development, from a single cell stage to a 16-somite stage.
%
We selected $120$ consecutive frames from this movie which capture 5.5~hours of epiboly (3.5--9~hours after fertilization).
%
In this experiment, embryos were immobilized for imaging so that the position and orientation remained fixed \cite{kane1996zebrafish}.
%
At the start of the time window, cells have divided 10--11 times and are accumulated in a cell mass above the yolk.
%
The cell mass is then compressed and the animal-vegetal axis of the embryo (vertical axis in \fig~\ref{fig:zebrafish}) shortens to form a spherical embryo shape by the end of the fourth hour of development.
%
Then, the yolk syncytial layer, which forms the boundary between the yolk and the cell mass, moves upward, forming a dome-shaped structure.
%
During this stage, the cells rearrange to form a uniform layer about four cells thick.
%
With time, this cell layer then spreads over across the yolk and expands toward the vegetal pole.
%
At the end of epiboly, the blastoderm completely engulfs the yolk.

@@

\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps \citep{singer2012vector}.
%
However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI).
%
Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set.
%
Averaging successive images removes some of the interembryo variability, so that sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.

@@

Although the theory of vector diffusion maps readily extends to three-dimensional data with full three-dimensional rotations, calculation of the pairwise alignments rapidly becomes computationally prohibitive for such applications, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle).
%
However, ordering data via diffusion maps is readily possible for pre-registered three-dimensional data.
%
Therefore, we can apply our methodology to three-dimensional data, such as a stack of confocal images taken at different depths within the sample, by first registering the data using an alternative method, and then ordering the registered data using diffusion maps.

@@

We then used diffusion maps to order the registered three-dimensional data.
%
\fig~\ref{fig:wing_disc}B shows selected images from the data set ordered by diffusion maps.
%
In the original data set, each disc was assigned to 6 time classes (72--73~hr, 76.5--77.5~hr, 79--80~hr, 89--90~hr, 100--101~hr, and 110.5--111.5~hr after fertilization); these times are indicated in \fig~\ref{fig:wing_disc}B.
%
In the ordered set, the size of wing disc grows, and the intensity of the Dad-GFP signal increases as a function of time.
%
The rank correlation coefficient based on the time class is 0.9436.
%
There are some registration errors in the early images, which are primarily due to some wing discs having extra tissue attached to them (such as the image in \fig~\ref{fig:wing_disc}A and the fourth image in \fig~\ref{fig:wing_disc}B).
%
Even with such obstructions, we can order the images with high accuracy and extract a stereotypical developmental trajectory, shown in \fig~\ref{fig:wing_disc}C, by averaging (see \SI).
%
We can now easily see the growth of the wing disc; however, averaging several images blurs the finer scale structures within the images
%
Thus, this result shows that the algorithm can successfully order a three-dimensional data set.

@@

%Already at this point, the rate of collection of imaging data has surpassed the rate of manual image analysis.
%
%This necessitates the development of automated methodologies to organize such large data sets.
%
%To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other.
%
Temporal ordering of large-scale data was done in the context of molecular profiling studies, in which data are vectors describing the expression levels of different mRNA \citep{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
%
At the same time, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \citep{yuan2014automated, surkova2008characterization, fowlkes2008quantitative}, or using some {\em a priori} knowledge of the relevant developmental processes \citep{dubuis2013accurate}.
%
%The task of image registration has been widely studied \citep{zitova2003image}, for applications such as face recognition \citep{rowley1998rotation}, medical image registration \citep{hajnal2010medical}, and texture classification \citep{greenspan1994rotation}.
%
In contrast to most of the existing registration approaches which rely on the knowledge of appropriate landmarks in the images \citep{ian1998statistical} (such as the eyes in face recognition applications \citep{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them relevant for a wide variety of applications.
%

@@

%
Vector diffusion maps allow us to automatically register images, an essential task for many applications.
%
Simultaneously, the algorithm provides us with parameters to describe each image.
%
In the examples presented here, we have focused on ordering the images in time using the first vector diffusion maps coordinate.
%
In general, we can recover several coordinates which concisely and comprehensively describe the data set.
%
This parametrization can then be used for typical data analysis tasks, such as outlier detection and model fitting.
%
Data which comes from several different genetic backgrounds can be clustered according to their spatial expression patterns, and clustering can also be used to separate images into different developmental stages.
%
Furthermore, images taken from different viewing directions can be analyzed, as the vector diffusion maps parametrization will organize the images according to the viewing angle \citep{singer2011viewing}.
%
Another direction for future work is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \citep{krzic2012multiview, ichikawa2014live, rubel2010coupling}.
%
Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.

@@

Images were subsampled, normalized, blurred, and centered prior to diffusion maps analysis to remove any variations due to the epxperimental and imaging framework.
%
Details about the preprocessing operations applied to each imging data set are given in \SI.

>>>>

Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
%
A schematic illustration of angular synchronization is shown in \fig~\ref{fig:schematics}A, where each image is represented as a vector, and the goal is to align the entire set of vectors given pairwise alignment measurements.
%
We first compute the angles needed to align pairs of vectors (or images), which in general requires no notion of a template function \citep{ahuja2007template, sonday2013noisy}.
%
In this work, we aligned pairs of images with respect to rotations by exhaustively searching over a discretized space of rotation angles to minimize the Euclidean distance between the pixels.
%
However, pairwise alignments can be computed by aligning appropriate image landmarks or features \citep{ian1998statistical}.
%
When the data are noisy, these pairwise measurements may by inaccurate, and so we utilize {\em all} pairwise measurements to align the set of images robustly.
%
Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with {\it all} pairwise measurements (see \SI); this is illustrated in \fig~\ref{fig:schematics}B.
%
In this schematic, registration via angular synchronization is trivial, as the pairwise measurements contain no noise.
%
However, the algorithm can register data sets even when many of the pairwise measurements are inaccurate \citep{singer2011angular}.

@@

To assess the robustness of the proposed methodology, we repeated this procedure with four additional data sets extracted from independent live imaging studies spanning the same developmental time period.
%
The results are shown in \fig~\ref{fig:drosophila_live_imaging}E.
%
The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than 95\%, indicating that our methodology can reproducibly order data of this type.
%
In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
%
At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments of the images, as well as the size of the data set \citep{singer2011angular}.
%
%Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
%%
%To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%%
%Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered.
%%
%\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%%
%As expected, the accuracy in the recovered ordering increases with the number of images.
%%
%At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
%%
%Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignment, and the high temporal resolution of our live imaging study allows us to order the images accurately.
%
%In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

@@

As another validation for the proposed methodology, we applied our algorithm to a time-lapse movie  of zebrafish embryogenesis.
%
We used a publicly available live imaging data of zebrafish embryogenesis (\texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov}, \citet{karlstrom1996flipbook}).
%
Taken with a differential interference contrast (DIC) microscope, the movie records the first $17$~hours of zebrafish development, from a single cell stage to a 16-somite stage.
%
We selected $120$ consecutive frames from this movie which capture 5.5~hours of epiboly (3.5--9~hours after fertilization).
%
In this experiment, embryos were immobilized for imaging so that the position and orientation remained fixed \citep{kane1996zebrafish}.
%
At the start of the time window, cells have divided 10--11 times and are accumulated in a cell mass above the yolk.
%
The cell mass is then compressed and the animal-vegetal axis of the embryo (vertical axis in \fig~\ref{fig:zebrafish}) shortens to form a spherical embryo shape by the end of the fourth hour of development.
%
Then, the yolk syncytial layer, which forms the boundary between the yolk and the cell mass, moves upward, forming a dome-shaped structure.
%
During this stage, the cells rearrange to form a uniform layer about four cells thick.
%
With time, this cell layer then spreads over across the yolk and expands toward the vegetal pole.
%
At the end of epiboly, the blastoderm completely engulfs the yolk.

@@

\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps.
%
However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI).
%
Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set.
%
Averaging successive images removes some of the interembryo variability, so that sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.

@@

In this section we show that the approach can readily be applied to three-dimensional data. 
%
We restrict ourselves to the case where an obvious fixed axis exists, so that only rotations of the three-dimensional data around this axis need be taken into account.  
%
This does not constitute an inherent limitation for vector diffusion maps. 
%
While for simplicity here we will not discuss the general  case, incorporating general 3D symmetries is possible \citep{arie2012global, wang2013exact, cucuringu2012eigenvector}.

%Although the theory of vector diffusion maps readily extends to three-dimensional data with full three-dimensional rotations, calculation of the pairwise alignments rapidly becomes computationally prohibitive for such applications, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle).
%
%However, ordering data via diffusion maps is readily possible for pre-registered three-dimensional data.
%
%Therefore, we can apply our methodology to three-dimensional data, such as a stack of confocal images taken at different depths within the sample, by first registering the data using an alternative method, and then ordering the registered data using diffusion maps.

@@

We then used diffusion maps to order the registered three-dimensional data.
%
\fig~\ref{fig:wing_disc}B shows selected images from the data set ordered by diffusion maps.
%
In the original data set, each disc was assigned to 6 time classes (72--73~hr, 76.5--77.5~hr, 79--80~hr, 89--90~hr, 100--101~hr, and 110.5--111.5~hr after fertilization); these times are indicated in \fig~\ref{fig:wing_disc}B.
%
In the ordered set, the size of wing disc grows, and the intensity of the Dad-GFP signal increases as a function of time.
%
The rank correlation coefficient based on the time class is 0.9436.
%
There are some registration errors in the early images, which are primarily due to some wing discs having extra tissue attached to them (such as the image in \fig~\ref{fig:wing_disc}A and the fourth image in \fig~\ref{fig:wing_disc}B).
%
Even with such obstructions, we can accurately order the images and extract a stereotypical developmental trajectory, shown in \fig~\ref{fig:wing_disc}C, by averaging (see \SI).
%
We can now easily see the growth of the wing disc, even though averaging somewhat blurs some finer scale structures.
%
%Thus, this result shows that the algorithm can successfully order a three-dimensional data set.

@@

%Already at this point, the rate of collection of imaging data has surpassed the rate of manual image analysis.
%
%This necessitates the development of automated methodologies to organize such large data sets.
%
%To the best of our knowledge, algorithmic approaches to these two tasks have been explored largely independently of each other.
%
Temporal ordering of large-scale data was done in the context of molecular profiling studies, in which data points are vectors describing the expression levels of different mRNA \citep{anavy2014blind, trapnell2014dynamics, gupta2008extracting}.
%
At the same time, temporal ordering of imaging data sets was done with a significant amount of human supervision and using registered images as a starting point \citep{yuan2014automated, surkova2008characterization, fowlkes2008quantitative}, or using some {\em a priori} knowledge of the relevant developmental processes \citep{dubuis2013accurate}.
%
%The task of image registration has been widely studied \citep{zitova2003image}, for applications such as face recognition \citep{rowley1998rotation}, medical image registration \citep{hajnal2010medical}, and texture classification \citep{greenspan1994rotation}.
%
In contrast to most of the existing registration approaches which rely on the knowledge of appropriate landmarks in the images \citep{ian1998statistical} (such as the eyes in face recognition applications \citep{zhao2003face}), algorithms based on angular synchronization can register images even in the absence of such information, making them relevant for a wide variety of applications.
%

@@

%
Vector diffusion maps allow us to automatically register images, an essential task for many applications.
%
Simultaneously, the algorithm provides us with parameters to describe each image.
%
In the examples presented here, we have focused on ordering the images in time using the first vector diffusion maps coordinate.
%
In general, we can recover several coordinates which concisely and comprehensively describe the data set.
%
This parametrization can then be used for typical data analysis tasks, such as outlier detection and model fitting.
%
Data which comes from several different genetic backgrounds can be clustered according to their spatial expression patterns, and clustering can also be used to separate images into different developmental stages.
%
Furthermore, images taken from different viewing directions can be analyzed, as the vector diffusion maps parametrization will organize the images according to the viewing angle \citep{singer2011viewing}.
%
Another direction for future work is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \citep{krzic2012multiview, ichikawa2014live, rubel2010coupling, dsilva2013nonlinear}.
%
Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.

@@

Images were subsampled, normalized, blurred, and centered prior to diffusion maps analysis to remove any variations due to the experimental and imaging framework.
%
Details about the preprocessing operations applied to each imging data set are given in \SI.
